The Spine of American Legal Practice

### 2016-06-29 article/article.Rmd
Kellen's changes to draft, shortens it quite a bit

At the opening of the first Nevada legislature in 1861, Territorial Governor James W. Nye, a former New York lawyer, instructed the assembly that they would have to forsake the Mormon statutes of Utah Territory from which Nevada had been carved. Those laws were ill adapted to "the mining interests" of the new territory, but "happily for us, a neighboring State whose interests are similar to ours, has established a code of laws" sufficiently attractive to "capital from abroad." That neighbor, of course, was California, and Nye urged that California's code of civil practice be enacted in Nevada, as far as it could "be made applicable."[^1] Territorial Senator William Morris Stewart, a famed silver-mining lawyer who would become the leading U.S. senator of Reconstruction, followed the instructions perhaps too well. Stewart literally cut and pasted the latest *Wood's Digest* of the California Practice Act into the session bill, crossing out *state* and *California* and substituting *territory* and *Nevada* where necessary. Stewart copied not just California's procedure code but also its method of codification, for California had in turn borrowed its code by modifying New York's.

@@

Nye wrote back to the assembly in disgust. The bill---of 715 sections---had reached him late the night before the legislative session was to close. Even in the few hours he had to read it, Nye counted "many errors in the enrolling of it, numbering probably more than three hundred." Some errors were severe. Nevada's Organic Act specified the jurisdiction of the district courts and justices of the peace, but the code overwrote these limitations by copying California's arrangements. Error-riddled and unconstitutional as the bill was, Nye believed a civil practice code was a "universal necessity and public need," concluding that "without some legislation in providing a code of procedure in civil cases, it may be a matter of grave question whether your courts would be able to fulfill the purpose of their creation."[^2] Nye signed the code into law.

Nothing like this "universal necessity" for a code of practice existed when Nye began his legal career in New York in the 1840s; rather, it was one of the central developments of America law during the nineteenth-century. By the end of the nineteenth-century, thirty-one American states and territories had adapted the text of a New York code of civil procedure first promulgated in 1848. The code became known as the Field Code, after its chief draftsman David Dudley Field, a famed Manhattan trial lawyer.[^3] When Field's code appeared in the Colorado assembly in 1876, a Denver newspaper wryly commented, "The scissors and paste-pot we had heretofore confidently believed were implements peculiar to the newspaper sanctum." To the editor, the cut-and-paste code was not just a curiosity of legislative history. Rather, the extensive copying of Field's Code threatened the foundations of American popular sovereignty. "The bill is a long one; the assembly has not the time to devote to it and to give it the scrutiny that a measure of such importance demands," he explained. Blind approval "would be an injustice to themselves and a greater wrong and injustice to the people who have a right to demand that their public servants legislate for the public good." The code "has been 'assimilated,' as we are informed, 'to the character and requirements of our people,' whatever that may mean," but the editor feared the legislation was the product of "men who have the welfare of the 'dear people' at their tongue's end always, but never in their hearts."[^4]

The migration of the Field Code was a central event in Anglo-American legal history, but no historian has traced the extensive borrowings of the Field text nor recognized the political furor that greeted the code outside New York.[^5] Every aspect of a civil justice system, from the rules granting access to courts, to lawyers, to remedies (whether damages, injunctions, or seizure of property) was covered by the code, making its New York-specific rules politically contentious both inside and outside the Empire State. As the Field Code migrated around the country, commentators in each jurisdiction raised the same complaint: how could legislation borrowed from another state represent the popular will and bests interests of *this* state?

Understanding the history of the Field Code requires not only attention to its political context but also a detailed examination of the substance of what was borrowed and what was revised in each jurisdiction. Exploring these borrowings is a daunting task, however. Procedure codes were long, technical documents, and although each jurisdiction copied large swaths of text, each also modified the text along the way, sometimes with a simple *Nevada* for *California*, sometimes with more foundational changes to the remedial system. Although Stewart's cut-and-paste code found its way into the archives, most draft legislation did not, and few codifiers explained in detail how they produced their texts. Traditional close reading or textual criticism of some 98,000 distinct sections of law comprising some 7.7 million words is simply not a feasible research task for a historian who wishes to track these borrowings.

Yet by turning to the digital analysis of texts, we have resolved this difficulty and tracked how states borrowed their codes of civil procedure from one another. Within the corpus of legislation, algorithmic analysis of texts can reverse engineer and visualize what the archive revealed in figure 1: which texts were borrowed, which modified, and how extensively. Our method works especially well for legal texts, for reasons we will explain. However, scholars can extend this method to other kinds of curated collection of texts, from church hymns to medical treatises.

The first contribution of this article is to demonstrate our methods as applied to a corpus of nineteenth-century state procedure codes. The second contribution is to integrate what we learned from the text analysis with the more conventional approaches of political and cultural history to explain why the migration of the Field Code mattered. On the national level the extent of legislative borrowing followed a pattern American historians have described as a "Greater Reconstruction" in which the former Confederate South and the Far West showed a remarkable kinship. Scholars have typically described Greater Reconstruction as a federal development, featuring the creation of national citizenship, a national economy, and a larger state apparatus centered in Washington, D.C. This article shows that Greater Reconstruction had its state-level dimensions as well. The uniform practice of law and adjudication of civil remedies was not structured by Washington mandates, however, but by the anxiety that New York financial capital would follow only New York civil remedies. At the more local level, our digital computations can trace modifications within code traditions, for instance, the ways western and midwestern codifiers altered New York's law to accommodate hardening conceptions of racial competencies in the civil courts.

## 1. The origin and migration of the Field Code

Until 1848, civil remedies and trial practice in New York were largely governed by common law traditions loosely categorized as "practice and pleadings."[^6] To understand how to file a civil claim or to enforce a judgment, a lawyer had to consult ad hoc statutes from colonial times to the present as well as precedents reported from cases litigated at common law and chancery. By the 1840s, enterprising practitioners had collated these materials into a half dozen marketable treatises, but these remained works of private opinion---no court was bound to agree with the treatise writers as to the weight, relevance, or proper interpretation of a legal statement.[^7] The common law was accordingly known as "unwritten law" despite the proliferation of published texts, because the common law was not precisely determined until a particular case drawn from the messy facts of life demanded resolution.[^8] Statutes, on the other hand, were "written law," prescribing or reforming the rules even before a case put the precise question in issue. Within the realm of written law, codes were the ultimate statutes.[^9]

@@

In New York, these codification debates came to a head at the 1846 constitutional convention, where "the conquerors took all," as the ambivalent Jacksonian James Fenimore Cooper complained.[^12] Law reformers abolished the court of chancery, made judges stand for popular election, and required the legislature to appoint a commission to reform the "practice and pleadings" of the civil courts.[^13]

@@

The other model came from England. Royal commissions had been employed since before the Revolution of 1688 to advise on a variety of matters. After the Revolution, royal commissions became disfavored while Parliament undertook many of the same functions through "committees of enquiry" to which only members of Parliament were appointed. By the nineteenth century, royal commissions again became common. Royal commissioners were usually not legislators. Although by the mid-nineteenth century royal commissions sometimes offered model statutes, the legislation they advised still had to pass through the normal politicking and drafting processes of Parliament.[^15]

After David Dudley Field and two other lawyers were appointed to the procedural commission, their reports made clear that they favored the French model but understood political realities would hold them to the English model. From 1847 to 1850, the commissioners made five reports to the legislature knowing they had no power to keep legislators from amending their code or even defeating it altogether. Each time they reminded legislators that "public opinion had issued its mandate in the most imposing form" of a constitutional decree. The constitution, together with the legislative act appointing the commission, "gave the commissioners instructions so precise, as to leave them no discretion, if they had desired it, \[and\] promised them therefore in advance, so long as they obeyed those instructions, the concurrence and co-operation of all departments of the government."[^16] Although the theory of codification made it a democratic enterprise, in practice Field---an ardent Jacksonian---insisted the democratic legislature ought to defer to the expertise of the commissioners.

@@

Yet the scope of the procedure Code included far more than the "mere machinery" of a lawsuit.[^18] The final draft of the code, printed in 1850, spanned nearly 800 pages of 1,885 regulations. The first third of the code covered constitutional topics, specifying the jurisdiction of all state courts and the duties of all state officers (and liabilities for violating those duties). The code deregulated attorney compensation, introducing novel structures of retainers and contingency fees.[^19] It created summary procedures meant to accelerate debt collection while simultaneously carving out "homestead" exemptions from the sheriff's reach.[^20] The code concluded by defining who could be an attorney, a juror, and a witness, drawing racial and gendered distinctions over who could speak in court.[^21]

Most important, the code defined all the remedies that a civil court could order---from money damages, to partition of property, to injunctive decrees and contempts---and made those remedies available in every lawsuit. In many cases, a legal right was indistinguishable from the remedy that secured that right: the right to possess a particular piece of property and the remedy that seized and delivered that property were, in effect, the same thing. Remedies were thus intimately connected to substantive law. For that reason, neither the French *code de procédure civile* (1806) nor Blackstone's and Bentham's writings conceived of remedies as purely procedural.[^22] By codifying remedies, Field invited continual expansion of the category of "procedure." As other states adopted the procedure code, they sometimes included other fields of law that seemed obviously "substantive" yet had such specific procedures or remedies that they were placed in a "code of procedure." Such fields included the law of wills, corporations, and mortgages.[^23] After all, argued procedural codifiers in Iowa, what did the famed Married Women's Property Acts offer besides procedural reform? These acts gave women *standing* to litigate in their own name and seek *remedies* in claims of property and contract, and they abolished mandatory rules of *joinder* (of husbands). Thus, one of the most significant changes to the law of property and domestic relations in the century went into the state's Code of Civil Procedure.[^24]

Despite Field's arguments that a mere procedure code was democratically unproblematic, his efforts were not entirely successful in New York. The commission submitted a draft of its main reforms in 1848, emphasizing that this first code was "but a report in part." New York's legislators enacted the partial code with little amendment, some legislators repeating Field's view that the constitution obligated them to accept the code.[^25] But when the commissioners submitted their final draft in 1849, the Assembly judiciary committee balked, directly disputing the commissioners' claims that procedure was merely the machinery of the law. They wrote that "provisions for rights and for the mode of pursuing remedies, insensibly run into each other," and observed that distinctions between substance and procedure were difficult to draw in practice. In fact, "the parts of a system of law are more dependent on each other, and run into and affect each other, infinitely more than any machine of human contrivance." They therefore suspected the commissioners' forthcoming code of criminal procedure would include all of the criminal law as well, "as they seem to understand practice and pleadings to include all the law upon a given subject." That being the case, the committee wondered whether they should "place in \[the commissioners\] a blind and implicit confidence that shall commit to their discretion the peace and property, the personal liberty and the lives of those who sent *us* here to make laws for them?"[^26]

![This map shows which states adopted codes of civil procedure based on the New York Field Code. The date shown is the date of the first enactment of a procedure code; most states subsequently revised their codes. Note that many southern states and western states came to adopt the Field Code during the period of Greater Reconstruction. By the end of the nineteenth-century, thirty-one jurisdictions (those displayed on the map, plus Alaska) had adopted a version of the Field Code. Data adapted from Charles McGuffey Hepburn, *The Historical Development of Code Pleading in America and England* (1897).](field-code-states-map.jpeg)

The code would encounter similar difficulties in each jurisdiction that adopted it. Even the shortest version of the Field Code was significantly longer than any other state statute before the Progressive legislation of the twentieth century. Unlike statutory compilations that sometimes took the name of a "code" but made no changes to existing law, the Field Code opened by abolishing the hallmarks of prior practice and instituting "hereafter" a new form of action with substantial revisions to basic matters of standing, remedy, and enforcement of civil judgments.[^27] In the states where it was imported, there was no getting around the fact that the code introduced much new law, yet legislators were unable to read, critique, and amend the code within the brief period of a legislative session. "It is folly to undertake to pass a code in a sixty day session," wrote the *Montana Post*, "and the best way would be for the Assembly to select one from a State or Territory which would come near meeting our wants, and slide it through with the fewest changes possible."[^28] Sliding the code through eased the problem of time but exacerbated the problem of local sovereignty. "To be governed by a foreign law, especially when that law is not preknown to the people whose conduct is to be regulated thereby ... is something repugnant to the idea of Democratic Republican government," complained the *The Miner's Express* in Iowa.[^29]

How, then, did states and territories achieve a politically acceptable balance between efficiency and sovereignty, borrowing law for sake of time but endowing it with popular legitimacy in each locale? Quite apart from the technicalities of legal practice, the American federation of civil government into (depending on the year) more than thirty or forty separate state jurisdictions makes it hard to describe a phenomenon that was truly national despite its state-centered enactments. It requires a sense of how much law was borrowed in each location and to what degree innovations were introduced. But precisely because these questions concern codes---texts that comprehensively and systematically cover a given subject---they are ideal sources for the techniques of digital history.

@@

To discover how the Field Code migrated to other jurisdictions, we compiled a corpus of relevant (or possibly relevant) laws, including separately bound codes of civil procedure as well as codes or regulations appearing as session laws and statutory compilations. The corpus comprises 135 statutes from the nineteenth century, which amounts to 7.7 million words organized into 98,000 regulations. It includes the initial enactment of every code of civil procedure, as well as procedure statutes and re-enacted codes from states reputed to have been legally influential, but it does not include every known statute of American procedural law. While a comprehensive project may be illuminating in its own ways, our specific question of how New York legislation influenced other states permits a more curated corpus.[^30]

Curating a corpus to answer a specific question is an important difference from the way that digital history often describes its work. Speaking in broad terms, digital historical work---like all historical work---can begin either with a set of sources or with a question. Computational digital history is often conceived of as beginning with very large datasets such as the Google Books or Hathi Trust corpora. These large corpora are sometimes called "big data"---though, it must be emphasized, almost never by digital historians who actually work with them---and are used as the basis of a "distant reading." We think that working with these large corpora is worthwhile because it is salutary for historians to have their research questions shaped by the broadest possible contexts. Nevertheless, it is not apparent that digital historians can readily move from these omnibus corpora to answering the specific research questions that animate various historical fields. In this article we demonstrate an alternative approach, which we might unimaginatively label "medium data." The amount of legislation that governed American civil practice is impressive, since every state amended and re-enacted procedure statutes nearly every decade. But while a corpus of procedural legislation requires some computational sophistication, it is far from needing the computational techniques derived for truly big data. We have gathered a large but narrowly constrained corpus centered on solving a well-defined research question. This corpus is large enough that digital historical methods provide results that a scholar could not obtain through traditional methods, but sufficiently circumscribed so as to directly address a discipline- and field-specific question.[^31]

@@

The most important step we took in processing the files was to split each section of the code into its own text file. Codes varied in how they were organized, but they all divided specific regulations into *sections* (or, on occasion, *articles*). Not only does the discursive form of these texts provide a handy organizational scheme for digital methods, but historically sections were also the way legislators borrowed their texts. Codifiers took their sources apart by sections, rearranging here, editing, drafting, and then re-combining there. Despite the fact that states differed widely on what topics they included in "civil procedure," sectioning the codes allowed us to assess similarity even among codes of quite different lengths and coverage. For instance, we know that California's 1851 code was derived from New York's 1850 code. (Stephen Johnson Field, David Dudley Field's brother, was the lawyer who imported New York's code into California.[^33]) But the New York code is over 150,000 words long, whereas California's code was just over 50,000 words long. Those disparate lengths mean that comparing all of the California code to all of the New York code is less meaningful than comparing each section in the California code to each section in the New York code, where matching sections will have a similar length.

@@

The aim, then, was to create a triangular matrix like the one above, but with approximately 98,000 rows and 98,000 columns, containing the similarity scores for each possible pair of sections. While this is easy enough to conceptualize, such a matrix is actually quite large, containing about 4.8 billion comparisons. This would take an unreasonable amount of computation time, and most of these comparisons would be unnecessary since each section has no relationship to most other sections. We therefore implemented the minhash/locality sensitive hashing algorithm to detect pairs of possible matches quickly.This algorithm works by extracting a set number of random tokens from each document to create a document signature; this subset makes it possible to approximate the Jaccard similarity of two documents. These signatures can then be used to identify each of the billions of possible pairs of sections as likely to be matches or not. By setting parameters to the algorithm, one can control how sensitive the algorithm is in drawing the line between genuine and spurious matches. We then measured the actual Jaccard score for only those candidate pairs.[^35]

@@

The New York Field Codes, especially the finished draft of 1850, were central to the entire network.[^39] New York gave rise to different regional traditions within the procedural network. Variations in the Field drafts meant that different states could borrow different versions of the Field Code. The 1850 New York code was the primary progenitor of several families of codes in California, Kentucky, Iowa, and Ohio, each of which in turn became major contributors to the law of neighboring states. The 1851 New York code became the progenitor of codes for Wisconsin, Florida, North Carolina, and South Carolina. Mississippi and Missouri borrowed from the New York 1849 code. While Field considered the 1850 version to be the definitive, ideal version of the code, all of the New York codes from 1848 to 1853 became models for other jurisdictions. In many cases, the commissions likely used whatever version of the code they had at hand. The Field Code was not a single volume on the shelf, but a series of volumes, any of which might be more accessible in different regions and in different years.

Even later New York codes can be considered a separate family. In 1876 a New York commission produced a new code attempting to consolidate all the statutory amendments subsequent to the 1851 Field Code. David Dudley Field was upset by the changes introduced in this revision, called the Throop Code. A count by a "friend" of Field's found that only three sentences of the Field Code had carried over word-for-word into the Throop Code. Nevertheless, many of the changes were purely formal, and Field speculated that these changes had been made only "so that \[the code\] should not appear to be the same thing as before."[^40] With respect to Mr. Field or his "friend," we found that the connection was between the Throop Code and the Field Code was somewhat stronger than he thought, and the Throop Code was indeed a descendant, if much revised, from the Field Code.

@@

In addition to the overview of the relationship between codes, we can also see more detail by visualizing the pattern of borrowings within each code. We have visualized the origin of each section of every code to see how each code was put together. To illustrate this, we will follow one branch of the Field Code network, beginning with the family started by California's 1850 and 1851 codes.

California's 1850 code, enacted in the period when California was entering the Union as a state, was borrowed almost entirely from New York's 1849 Field Code. The compiler Elisha Crosby did lift one significant portion regulating when a judge could order a new trial from the mixed civilian/common law code of Louisiana . New trials were not provided for in the New York Code until the finished draft in 1850. Most of the sections that were not borrowed, as with many of the codes, have to do with parts that describe the system of courts or provide sample forms peculiar to that state.

@@

## 4. Procedure codes and capitalism

>>>>

At the opening of the first Nevada legislature in 1861, Territorial Governor James W. Nye, a former New York lawyer, instructed the assembly that they would have to forsake their inherited Mormon statutes that were ill adapted to "the mining interests" of the new territory. "Happily for us, a neighboring State whose interests are similar to ours, has established a code of laws" sufficiently attractive to "capital from abroad." That neighbor was California, and Nye urged that California's "Practice Code" be enacted in Nevada, as far as it could "be made applicable."[^1] Territorial Senator William Morris Stewart, a famed mining lawyer who would lead the U.S. Senate during Reconstruction, followed the instructions perhaps too well. Stewart literally cut and pasted the latest *Wood's Digest* of the California Practice Act into the session bill, crossing out *state* and *California* and substituting *territory* and *Nevada* where necessary. Stewart copied not just California's procedure code but also its method of codification, for California had in turn borrowed its code by modifying New York's.

@@

Nye wrote back to the assembly in disgust. The bill---of 715 sections---had reached him late the night before the legislative session was to close. Even in the few hours he had to read it, Nye counted "many errors in the enrolling of it, numbering probably more than three hundred." Some errors were severe. The code overwrote the specific jurisdictional boundaries of Nevada’s Organic Act by copying California’s arrangements. Error-riddled and unconstitutional as the bill was, Nye believed a civil practice code was a "universal necessity and public need," doubting "whether your courts would be able to fulfill the purpose of their creation" without one.[^2] Nye signed the code into law.

Nothing like this "universal necessity" existed when Nye began his legal career in New York in the 1840s; rather, it was one of the central developments of America law after the mid-nineteenth century. By 1900, thirty-one American states and territories had adapted the text of a New York code of civil procedure first promulgated in 1848. The code became known as the Field Code, after its chief draftsman David Dudley Field, a Manhattan trial lawyer.[^3] When Field's code appeared in the Colorado assembly in 1876, a Denver newspaper wryly commented, "The scissors and paste-pot we had heretofore confidently believed were implements peculiar to the newspaper sanctum." To the editor, the cut-and-paste code was not just a curiosity of legislative history. Rather, the extensive copying of Field's Code threatened the foundations of American popular sovereignty. "The bill is a long one; the assembly has not the time to devote to it and to give it the scrutiny that a measure of such importance demands.” Blind approval "would be an injustice to themselves and a greater wrong and injustice to the people who have a right to demand that their public servants legislate for the public good." The code "has been 'assimilated,' as we are informed, 'to the character and requirements of our people,' whatever that may mean," but the editor feared the legislation was the product of "men who have the welfare of the 'dear people' at their tongue's end always, but never in their hearts."[^4]

The migration of the Field Code was a central event in Anglo-American legal history, but no historian has traced the extensive borrowings of the Field text nor recognized the political furor that greeted the code outside New York.[^5] Every aspect of a civil justice system, from the rules granting access to courts, to lawyers, to remedies (whether damages, injunctions, or seizure of property) was covered by the code, making its New York-specific rules politically contentious both inside and outside the Empire State. As the Field Code migrated around the country, commentators in each jurisdiction raised the same complaint: how could legislation borrowed from another state represent the popular will and best interests of *this* state?

Understanding the history of the Field Code requires not only attention to its political context but also a detailed examination of the substance of what was borrowed and what was revised in each jurisdiction. Exploring these borrowings is a daunting task, however. Procedure codes were long, technical documents, and although each jurisdiction copied large swaths of text, each also modified the text along the way, sometimes with a simple *Nevada* for *California*, sometimes with more foundational changes to civil remedies. Although Stewart's cut-and-paste code found its way into the archives, most draft legislation did not, and few codifiers explained in detail how they produced their texts. Traditional close reading or textual criticism of some 98,000 distinct sections of law comprising 7.7 million words is simply not a feasible research task for a historian who wishes to track these borrowings.

Yet by turning to the digital analysis of texts, we have resolved this difficulty and tracked how states borrowed their codes of civil practice from one another. Within the corpus of legislation, algorithmic analysis of texts can reverse engineer and visualize what the archive revealed in figure 1: which texts were borrowed, which modified, and how extensively. Our method works especially well for legal texts, for reasons we will explain. However, scholars can extend this method to other curated collections of texts, from church hymns to medical treatises.

The first contribution of this article is to demonstrate our methods as applied to a corpus of nineteenth-century civil procedure codes. The second contribution is to integrate what we learned from the text analysis with the more conventional approaches of political and cultural history to explain why the migration of the Field Code mattered. On the national level the extent of legislative borrowing followed a pattern American historians have described as a "Greater Reconstruction" in which the former Confederate South and the Far West showed a remarkable kinship. Scholars have typically described Greater Reconstruction as a federal development, featuring the creation of national citizenship, a national economy, and a larger federal apparatus centered in Washington, D.C. This article shows that Greater Reconstruction had its state-level dimensions as well. The uniform practice of law and adjudication of civil remedies was not structured by Washington mandates, however, but by the anxiety that New York financial capital would follow only New York civil remedies. At the more local level, our digital computations can trace modifications within code traditions, for instance, the ways western and midwestern codifiers altered New York's law to accommodate hardening conceptions of racial competencies in the civil courts.

## 1. The origin and political controversies of the Field Code

Until 1848, civil remedies and trial practice in New York were largely governed by common law traditions loosely categorized as "practice and pleadings."[^6] To understand how to file a civil claim or to enforce a judgment, a lawyer had to consult ad hoc statutes from colonial times to the present as well as precedents reported from cases litigated at common law and chancery. By the 1840s, enterprising practitioners had collated these materials into a half dozen marketable treatises, but these remained works of private opinion---no court was bound to agree with the treatise writers as to the weight, relevance, or proper interpretation of a legal statement.[^7] The common law was accordingly known as "unwritten law" despite the proliferation of published texts, because the common law was not precisely determined until a particular case demanded resolution.[^8] Statutes, on the other hand, were "written law," prescribing or reforming the rules even before a case put the precise question in issue. Within the realm of written law, codes were the ultimate statutes.[^9]

@@

In New York, these codification debates came to a head at the 1846 constitutional convention, where "the conquerors took all," as the ambivalent Jacksonian James Fenimore Cooper complained.[^12] Law reformers abolished the court of chancery, made judges stand for popular election, and required the legislature to appoint commissioners to codify the law and reform the "practice and pleadings" of the civil courts.[^13]

@@

The other model came from England. Royal commissions had been employed since before the Revolution of 1688 to advise on a variety of matters. Although by the mid-nineteenth century royal commissions sometimes offered model statutes, Parliament maintained exclusive legislative prerogatives, forcing any commissioner-proposed legislation to pass through the normal politicking and drafting processes of Parliament.[^15]

After David Dudley Field and two other lawyers were appointed to the procedural commission, their reports made clear that they favored the French model but understood political realities would hold them to the English model. From 1847 to 1850, the commissioners made five reports to the legislature knowing they had no power to keep legislators from amending their code or even defeating it altogether. Each time they reminded legislators that "public opinion had issued its mandate in the most imposing form" of a constitutional decree. The constitution, together with the legislative act appointing the commission, "gave the commissioners instructions so precise, as to leave them no discretion, if they had desired it, \[and\] promised them therefore in advance, so long as they obeyed those instructions, the concurrence and co-operation of all departments of the government."[^16] Although the theory of codification made it a democratic enterprise, in practice Jacksonians like Field insisted the democratic legislature ought to defer to the expertise of the commissioners.

@@

Yet the code's scope of "procedure" included far more than the "mere machinery" of a lawsuit.[^18] The final draft of the code, printed in 1850, spanned nearly 800 pages of 1,885 regulations. The first third of the code covered constitutional topics, specifying the jurisdiction of all state courts and the duties of all state officers (and liabilities for violating those duties). The code deregulated attorney compensation, introducing novel structures of retainers and contingency fees.[^19] It created summary procedures meant to accelerate debt collection while simultaneously carving out "homestead" exemptions from the sheriff's reach.[^20] The code concluded by defining who could be an attorney, a juror, and a witness, drawing racial and gendered distinctions over who could speak in court.[^21]

Most important, the code defined all the remedies that a civil court could order---from money damages, to partition of property, to injunctive decrees and contempts---and made those remedies available in every lawsuit. In many cases, a legal right was indistinguishable from the remedy that secured that right: the right to possess a particular piece of property and the remedy that seized and delivered that property were, in effect, the same thing. Remedies were thus intimately connected to substantive law. For that reason, neither the French *code de procédure civile* (1806) nor Blackstone or Bentham's writings conceived of remedies as purely procedural.[^22] By codifying remedies, Field invited continual expansion of the category of "procedure." As other states adopted the procedure code, they sometimes included other fields of law that seemed obviously "substantive" yet had such specific procedures or remedies that they were placed in a "code of procedure." Such fields included the law of wills, corporations, and mortgages.[^23] And after all, argued procedural codifiers in Iowa, what did the famed Married Women's Property Acts offer besides procedural reform? These acts gave women *standing* to litigate in their own name and seek *remedies* in claims of property and contract, and they abolished mandatory rules of *joinder* (of husbands). Thus, one of the most significant changes to the law of property and domestic relations in the century went into the state's Code of Civil Procedure.[^24]

Despite Field's arguments that a mere procedure code was democratically unproblematic, his efforts were not entirely successful in New York. The commission submitted a draft of its main reforms in 1848, emphasizing that this first code was "but a report in part." New York's legislators enacted the partial code with little amendment, some legislators repeating Field's view that the constitution obligated them to accept the code.[^25] But when the commissioners submitted an extended draft in 1849, the Assembly judiciary committee balked, directly disputing the commissioners' claims that procedure was merely the machinery of the law. The "provisions for rights and for the mode of pursuing remedies, insensibly run into each other," the committee reported, complicating legal practice "infinitely more than any machine of human contrivance." They therefore suspected the commissioners' forthcoming code of criminal procedure would include all of the criminal law as well, "as they seem to understand practice and pleadings to include all the law upon a given subject." That being the case, the committee wondered whether they should "place in \[the commissioners\] a blind and implicit confidence that shall commit to their discretion the peace and property, the personal liberty and the lives of those who sent *us* here to make laws for them?"[^26]

![This map shows which states adopted codes of civil procedure based on the New York Field Code. The date shown is the date of the first enactment of a procedure code; most states subsequently revised their codes. Note that many southern states and western states came to adopt the Field Code during the Civil War and Reconstruction. By the end of the nineteenth-century, thirty-one jurisdictions (those displayed on the map, plus Alaska) had adopted a version of the Field Code. Data adapted from Charles McGuffey Hepburn, *The Historical Development of Code Pleading in America and England* (1897).](field-code-states-map.jpeg)

The code would encounter similar difficulties in each jurisdiction that adopted it. Even the shortest version of the Field Code was significantly longer than any other state statute before the Progressive legislation of the twentieth century. Unlike statutory compilations that sometimes took the name of a "code" but made no changes to existing law, the Field Code opened by abolishing the hallmarks of prior practice and instituting "hereafter" a new form of action with substantial revisions to basic matters of civil remedies.[^27] In the states where it was imported, there was no getting around the fact that the code introduced much new law, yet legislators were unable to read, critique, and amend the code within the brief period of a legislative session. "It is folly to undertake to pass a code in a sixty day session," wrote the *Montana Post*, "and the best way would be for the Assembly to select one from a State or Territory which would come near meeting our wants, and slide it through with the fewest changes possible."[^28] Sliding the code through eased the problem of time but exacerbated the problem of local sovereignty. "To be governed by a foreign law, especially when that law is not preknown to the people whose conduct is to be regulated thereby ... is something repugnant to the idea of Democratic Republican government," complained the *The Miner's Express* in Iowa.[^29]

How, then, did states and territories achieve a politically acceptable balance between efficiency and sovereignty, borrowing law for sake of time but endowing it with popular legitimacy in each locale? Quite apart from the technicalities of legal practice, the American federation of civil government into (depending on the year) more than thirty or forty separate jurisdictions makes it hard to describe a phenomenon that was truly national despite its state-centered enactments. It requires a sense of how much law was borrowed in each location and to what degree innovations were introduced. But precisely because these questions concern codes---texts that comprehensively and systematically cover a given subject---they are ideal sources for the techniques of digital history.

@@

To discover how the Field Code migrated to other jurisdictions, we compiled a corpus of potentially relevant laws, including separately bound codes of civil procedure as well as codes or statutes appearing within session laws and statutory compilations from around the Atlantic world. The corpus comprises 135 statutes from the nineteenth century, which amounts to 7.7 million words organized into 98,000 regulations. It includes the initial enactment of every U.S. code of civil procedure, as well as procedure statutes and re-enacted codes from jurisdictions reputed to have been legally influential, including French and British codes. The corpus does not include every nineteenth-century statute of procedural law. While a comprehensive project may be illuminating in its own ways, our specific question of how New York legislation influenced other American jurisdictions permits a more curated corpus.[^30]

Curating a corpus to answer a specific question is an important difference from the way that digital history often describes its work. Speaking in broad terms, digital historical work---like all historical work---can begin either with a set of sources or with a question. Computational digital history is often conceived of as beginning with sources, particularly with large datasets such as the Google Books or Hathi Trust corpora. These large corpora are sometimes called "big data"---though, it must be emphasized, almost never by digital historians who actually work with them---on which "distant reading" can be practiced. While it is salutary for historians to have their research questions shaped by the broadest possible contexts, it is not apparent that digital historians can readily move from these omnibus corpora to answering the specific research questions that animate various historical fields. In this article we demonstrate an alternative approach, which we might unimaginatively label "medium data." The amount of legislation that governed American civil practice is impressive, since every state amended and re-enacted procedure statutes nearly every decade. But while a corpus of procedural legislation requires some computational sophistication, the techniques are far less complex than those derived for truly big data. We have gathered a large but narrowly constrained corpus centered on solving a well-defined research question. This corpus is large enough that digital historical methods provide results that a scholar could not obtain through traditional methods, but sufficiently circumscribed so as to directly address a discipline- and field-specific question.[^31]

@@

The most important step we took in processing the files was to split each section of the code into its own text file. Codes varied in how they were organized, but they all divided specific regulations into *sections* (or, on occasion, *articles*). Not only does the discursive form of these texts provide a handy organizational scheme for digital methods, but historically sections were also the way legislators borrowed their texts. Codifiers took their sources apart by sections, rearranging here, editing, drafting, and then re-combining there. Despite the fact that states differed widely on what topics they included in "civil procedure," sectioning the codes allowed us to assess similarity even among codes of quite different lengths and coverage. For instance, we know that California's 1851 code was derived from New York's 1850 code. (Stephen J. Field, David Dudley's brother, was the lawyer who imported New York's code into California.[^33]) But the New York code is over 150,000 words long, whereas California's code was just over 50,000 words long. Those disparate lengths mean that comparing all of the California code to all of the New York code is less meaningful than comparing each section in the California code to each section in the New York code, where matching sections will have a similar length.

@@

The aim, then, was to create a triangular matrix like the one above, but with approximately 98,000 rows and 98,000 columns, containing the similarity scores for each possible pair of sections. While this is easy enough to conceptualize, such a matrix is actually quite large, containing about 4.8 billion comparisons. This would take an unreasonable amount of computation time, and most of these comparisons would be unnecessary since each section has no relationship to most other sections. We therefore implemented the minhash/locality sensitive hashing algorithm to detect pairs of possible matches quickly. Instead of comparing all tokens to one another, this algorithm samples tokens from each document to find probable matches, and then Jaccard scores can be calculated for only those probable matches (that is, many of the needless calculations that produce scores of zero get cut out).[^35]

@@

The New York Field Codes, especially the finished draft of 1850, were central to the entire network.[^39] New York gave rise to different regional traditions within the procedural network. Variations in the Field drafts meant that different states could borrow different versions of the Field Code. Field’s 1850 draft---never actually enacted in New York---was the primary progenitor of several families of codes in California, Kentucky, Iowa, and Ohio, each of which in turn became major contributors to the law of neighboring states. The 1851 New York code---a small revision to the original 1848 code---became the progenitor of codes for Wisconsin, Florida, North Carolina, and South Carolina. While Field considered the 1850 version to be the definitive, ideal version of the code, all of the New York codes from 1848 to 1853 became models for other jurisdictions. In many cases, the commissions likely used whatever version of the code they had at hand. The Field Code was not a single volume on the shelf, but a series of drafts, any of which might be more accessible in different regions and in different years.

Even later New York codes can be considered a separate family. In 1876 a New York commission produced a new code attempting to consolidate all the case law and statutory amendments subsequent to the 1851 Field Code. David Dudley Field was upset by the changes introduced in this revision. A count by a "friend" of Field's found that only three sentences of the Field Code had carried over word-for-word into the latest edition. With respect to Mr. Field or his "friend," we found that the connection between the codes somewhat stronger than he thought, although his conclusion that, textually, the 1876 code did "not appear to be the same thing as before," remains sound.[^40]

@@

In addition to the overview of the relationship between codes, we can also see more detail by visualizing the pattern of borrowings within each code. To illustrate this, we will follow one branch of the Field Code network, beginning with the family started by California's 1850 and 1851 codes.

California's 1850 code, enacted in the period when California was entering the Union as a state, was borrowed almost entirely from New York's 1849 Field Code. The compiler Elisha Crosby did lift one portion from the mixed civilian/common law code of Louisiana, the rules for ordering a new trial to revisit an earlier jury’s verdict. New trials were not provided for in the New York Code until the finished draft in 1850. Most of the sections that were not borrowed, as with many of the codes, have to do with parts that describe the system of courts or provide sample forms of pleading or sheriff’s writs that were peculiar to each state.

@@

## 4. Procedure codes and American capitalism

The near uniformity of creditors' remedies in Field Code states, as demonstrated by clustering, points out the close link between the rise of modern American procedure and modern American capitalism. Westerners and southerners frequently commented on the seeming imperialism of the New York code and its connections to New York capital, but one must turn to the technical debates over procedure to find these anxieties. Twelve of the states and territories that copied the Field Code most closely did so during the Civil War and Reconstruction era---four states in the former Confederacy and eight jurisdictions in the Far West.[^46] As with other areas of postbellum history, it turns out one may learn a lot by holding the postbellum American South and American West together.
